Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-21 blinded-paper.tex
add submit archives

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

  \author{Juan Carlos Castillo}
            \affil{%
                  Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }

@@

This document was last modified at 2025-01-19 22:27:33

and it was last rendered at 2025-01-19 22:27:33

@@

Empirical cross-sectional evidence from Chile reveals two significant
trends: low public support for the perception of meritocracy and limited
justification for unequal access to social services such as education
and health. For instance, Castillo, Madero-Cabib, and Salamovich (2013)
analyzed data from the International Social Survey Programme (ISSP) in
1999 and 2009, highlighting consistently low support for inequality
based on resource allocation or individual merit. Similarly, Castillo,
Miranda, and Carrasco (2012) demonstrated widespread skepticism among
Chileans regarding the meritocratic system, with many disagreeing with
the notion that rewards are fairly distributed based on talent and
effort. In line with this evidence, we propose that:

@@

A third group of variables is included in the analysis to assess the
effect of elements related to the pandemic context. These variables were
only included in the 2021 wave, so each individual's responses were
replicated for the remaining years. The first variable addresses the
belief regarding whether, in the context of the pandemic, it is more
important to give priority to economic activity than to the health of
the population, measured on a Likert scale from (1) ``strongly
disagree'' to (2) ``strongly agree''. The second variable is a dummy
(No/Yes) that captures whether individuals received state benefits
during the pandemic (e.g., Food Box, Emergency Family Income,
Contribution for the Middle Class, or others).

@@

\caption{\label{fig-correlation}Correlation matrix of the main variables
for the first wave (2016)}

@@

~

Model 1

Model 2

Model 3

Model 4

Model 5

Model 6

Model 7

Model 8

Intercept

1.847***

1.853***

1.920***

1.776***

1.641***

1.563***

2.174***

2.183***

~

(0.026)

(0.038)

(0.048)

(0.064)

(0.068)

(0.069)

(0.086)

(0.120)

Wave (Ref.= Wave 2016)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Wave 2017

-0.167***

~

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~

~~~~~Wave 2018

-0.059*

~

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~

~~~~~Wave 2019

-0.030

~

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~

~~~~~Wave 2022

0.204***

~

~

~

~

~

~

~

~

(0.030)

~

~

~

~

~

~

~

~~~~~Wave 2023

0.233***

~

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~

Wave

~

-0.073***

-0.073***

-0.073***

-0.076***

-0.076***

-0.052**

-0.052**

~

~

(0.020)

(0.020)

(0.020)

(0.020)

(0.020)

(0.020)

(0.020)

Wave\^{}2

~

0.016***

0.016***

0.016***

0.016***

0.016***

0.013***

0.013***

~

~

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

Covid: receive state benefits

~

~

-0.095*

-0.097*

-0.100*

-0.102*

-0.102**

-0.093*

~

~

~

(0.041)

(0.041)

(0.040)

(0.040)

(0.039)

(0.041)

Covid: economy more important than health

~

~

~

0.065***

0.060**

0.055**

0.049**

0.038*

~

~

~

~

(0.019)

(0.019)

(0.019)

(0.019)

(0.019)

Merit: Talent

~

~

~

~

0.055***

-0.002

-0.003

-0.005

~

~

~

~

~

(0.010)

(0.012)

(0.012)

(0.012)

Merit: Effort

~

~

~

~

~

0.096***

0.084***

0.083***

~

~

~

~

~

~

(0.013)

(0.013)

(0.013)

Inequality perception

~

~

~

~

~

~

-0.135***

-0.134***

~

~

~

~

~

~

~

(0.012)

(0.012)

Universitary (Ref.= Less than universitary)

~

~

~

~

~

~

~

-0.048

~

~

~

~

~

~

~

~

(0.046)

Household income

~

~

~

~

~

~

~

0.001*

~

~

~

~

~

~

~

~

(0.000)

Controls

No

No

No

No

No

No

No

Yes

BIC

29667.072

29670.297

29678.484

29682.436

29668.980

29626.960

29510.881

29626.312

Num. obs.

7576

7576

7576

7576

7576

7576

7576

7576

Num. groups: Individuals

1322

1322

1322

1322

1322

1322

1322

1322

Var: Individuals (Intercept)

0.185

0.184

0.183

0.179

0.174

0.170

0.164

0.159

Var: Residual

0.497

0.500

0.500

0.500

0.499

0.496

0.488

0.488

*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
Note: Model 8 is controlled by age, gender, political position, and
subjective social status.

@@

~

Model 8

Model 9

Model 10

Intercept

2.18***

2.03***

2.09***

~

(0.12)

(0.13)

(0.13)

Wave

-0.05**

-0.05*

-0.05**

~

(0.02)

(0.02)

(0.02)

Wave\^{}2

0.01***

0.01***

0.01***

~

(0.00)

(0.00)

(0.00)

Covid: receive state benefits

-0.09*

0.11

0.03

~

(0.04)

(0.07)

(0.07)

Merit: Talent

-0.00

0.05*

-0.01

~

(0.01)

(0.02)

(0.01)

Merit: Effort

0.08***

0.08***

0.12***

~

(0.01)

(0.01)

(0.02)

Receive state benefits * Merit: Talent

~

-0.07**

~

~

~

(0.02)

~

Receive state benefits * Merit: Effort

~

~

-0.05*

~

~

~

(0.02)

BIC

29626.31

29630.35

29636.59

Num. obs.

7576

7576

7576

Num. groups: Individuals

1322

1322

1322

Var: Individuals (Intercept)

0.16

0.16

0.16

Var: Residual

0.49

0.49

0.49

\item *

** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
Note: All the models are controlled by educational level, income
quintile, subjective social status, age, gender, political position, and
inequality perception

@@

\section*{(APPENDIX) Appendix}\label{appendix-appendix}
\addcontentsline{toc}{section}{(APPENDIX) Appendix}

\appendix \section{Appendix}

@@

~

Model 1

Model 2

Model 3

Model 4

Model 5

Model 6

Model 7

Intercept

1.847***

1.853***

1.873***

1.741***

1.664***

2.276***

2.277***

~

(0.026)

(0.038)

(0.090)

(0.092)

(0.092)

(0.105)

(0.138)

Wave (Ref.= Wave 2016)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Wave 2017

-0.167***

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~~~~~Wave 2018

-0.059*

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~~~~~Wave 2019

-0.030

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

~~~~~Wave 2022

0.204***

~

~

~

~

~

~

~

(0.030)

~

~

~

~

~

~

~~~~~Wave 2023

0.233***

~

~

~

~

~

~

~

(0.029)

~

~

~

~

~

~

Wave

~

-0.073***

-0.073***

-0.076***

-0.076***

-0.052**

-0.052**

~

~

(0.020)

(0.020)

(0.020)

(0.020)

(0.020)

(0.020)

Wave\^{}2

~

0.016***

0.016***

0.016***

0.016***

0.013***

0.013***

~

~

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

Covid: receive state benefits

~

~

-0.097*

-0.100*

-0.102*

-0.102**

-0.093*

~

~

~

(0.041)

(0.040)

(0.040)

(0.039)

(0.041)

Covid: economy more important than health

~

~

0.065***

0.060**

0.055**

0.049**

0.038*

~

~

~

(0.019)

(0.019)

(0.019)

(0.019)

(0.019)

Merit: Talent

~

~

~

0.055***

-0.002

-0.003

-0.005

~

~

~

~

(0.010)

(0.012)

(0.012)

(0.012)

Merit: Effort

~

~

~

~

0.096***

0.084***

0.083***

~

~

~

~

~

(0.013)

(0.013)

(0.013)

Inequality perception

~

~

~

~

~

-0.135***

-0.134***

~

~

~

~

~

~

(0.012)

(0.012)

Universitary (Ref.= Less than universitary)

~

~

~

~

~

~

-0.048

~

~

~

~

~

~

~

(0.046)

Household Income

~

~

~

~

~

~

0.001*

~

~

~

~

~

~

~

(0.000)

Subjective Social Status

~

~

~

~

~

~

-0.005

~

~

~

~

~

~

~

(0.012)

Age (Ref.= 18-29)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Age 30-49

~

~

~

~

~

~

-0.033

~

~

~

~

~

~

~

(0.051)

~~~~~Age 50-64

~

~

~

~

~

~

0.009

~

~

~

~

~

~

~

(0.053)

~~~~~Age 65 or more

~

~

~

~

~

~

0.123

~

~

~

~

~

~

~

(0.066)

Gender (Ref. Male)

~

~

~

~

~

~

-0.082*

~

~

~

~

~

~

~

(0.037)

Pol. pos (Ref.= Left)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Center

~

~

~

~

~

~

0.095

~

~

~

~

~

~

~

(0.051)

~~~~~Right

~

~

~

~

~

~

0.220***

~

~

~

~

~

~

~

(0.059)

~~~~~Does not identify

~

~

~

~

~

~

0.085

~

~

~

~

~

~

~

(0.047)

BIC

29667.072

29670.297

29682.436

29668.980

29626.960

29510.881

29626.312

Num. obs.

7576

7576

7576

7576

7576

7576

7576

Num. groups: Individuals

1322

1322

1322

1322

1322

1322

1322

Var: Individuals (Intercept)

0.185

0.184

0.179

0.174

0.170

0.164

0.159

Var: Residual

0.497

0.500

0.500

0.499

0.496

0.488

0.488

*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

  \author{ }

@@

@@

Empirical cross-sectional evidence from Chile reveals two significant
trends: low perceived meritocracy and limited justification for unequal
access to social services such as education and health. For instance,
Castillo, Madero-Cabib, and Salamovich (2013) analyzed data from the
International Social Survey Programme (ISSP) in 1999 and 2009,
highlighting consistently low support for inequality based on resource
allocation or individual merit. Similarly, Castillo, Miranda, and
Carrasco (2012) demonstrated widespread skepticism among Chileans
regarding the meritocratic system, with many disagreeing with the notion
that rewards are fairly distributed based on talent and effort. In line
with this evidence, we propose that:

@@

A third group of variables is included in the analysis to assess the
effect of elements related to the pandemic context. These variables were
only included in the 2021 wave, so each individual's responses were
replicated for the remaining years. The first variable addresses the
belief regarding whether, in the context of the pandemic, it is more
important to give priority to economic activity than to the health of
the population, measured on a Likert scale from (1) ``strongly
disagree'' to (5) ``strongly agree''. The second variable is a dummy
(No/Yes) that captures whether individuals received state benefits
during the pandemic (e.g., Food Box, Emergency Family Income,
Contribution for the Middle Class, or others).

@@

\caption{\label{fig-correlation}Correlation matrix of the main variables
for the last wave (2023)}

@@

\begin{center}
\scalebox{0.765}{
\begin{tabular}{l c c c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
\hline
Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.920^{***}$  & $1.776^{***}$  & $1.641^{***}$  & $1.563^{***}$  & $2.174^{***}$  & $2.183^{***}$  \\
                                            & $(0.026)$      & $(0.038)$      & $(0.048)$      & $(0.064)$      & $(0.068)$      & $(0.069)$      & $(0.086)$      & $(0.120)$      \\
Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                &                \\
                                            &                &                &                &                &                &                &                &                \\
\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                &                \\
\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                &                \\
\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                &                \\
\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                &                \\
                                            & $(0.030)$      &                &                &                &                &                &                &                \\
\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                &                \\
Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
Covid: receive state benefits               &                &                & $-0.095^{*}$   & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
                                            &                &                & $(0.041)$      & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
Covid: economy more important than health   &                &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
                                            &                &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
Merit: Talent                               &                &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
                                            &                &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Merit: Effort                               &                &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
                                            &                &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
Inequality perception                       &                &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
                                            &                &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
Universitary (Ref.= Less than universitary) &                &                &                &                &                &                &                & $-0.048$       \\
                                            &                &                &                &                &                &                &                & $(0.046)$      \\
Household income                            &                &                &                &                &                &                &                & $0.001^{*}$    \\
                                            &                &                &                &                &                &                &                & $(0.000)$      \\
\hline
Controls                                    & No             & No             & No             & No             & No             & No             & No             & Yes            \\
BIC                                         & $29667.072$    & $29670.297$    & $29678.484$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.183$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
\hline
\multicolumn{9}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.

Note: Model 8 is controlled by age, gender, political position, and subjective social status.}}
\end{tabular}
}
\label{table:coefficients}
\end{center}

@@

\begin{center}
\scalebox{0.765}{
\begin{tabular}{l c c c}
\hline
 & Model 8 & Model 9 & Model 10 \\
\hline
Intercept                              & $2.18^{***}$ & $2.03^{***}$ & $2.09^{***}$ \\
                                       & $(0.12)$     & $(0.13)$     & $(0.13)$     \\
Wave                                   & $-0.05^{**}$ & $-0.05^{*}$  & $-0.05^{**}$ \\
                                       & $(0.02)$     & $(0.02)$     & $(0.02)$     \\
Wave$^2$                               & $0.01^{***}$ & $0.01^{***}$ & $0.01^{***}$ \\
                                       & $(0.00)$     & $(0.00)$     & $(0.00)$     \\
Covid: receive state benefits          & $-0.09^{*}$  & $0.11$       & $0.03$       \\
                                       & $(0.04)$     & $(0.07)$     & $(0.07)$     \\
Merit: Talent                          & $-0.00$      & $0.05^{*}$   & $-0.01$      \\
                                       & $(0.01)$     & $(0.02)$     & $(0.01)$     \\
Merit: Effort                          & $0.08^{***}$ & $0.08^{***}$ & $0.12^{***}$ \\
                                       & $(0.01)$     & $(0.01)$     & $(0.02)$     \\
Receive state benefits * Merit: Talent &              & $-0.07^{**}$ &              \\
                                       &              & $(0.02)$     &              \\
Receive state benefits * Merit: Effort &              &              & $-0.05^{*}$  \\
                                       &              &              & $(0.02)$     \\
\hline
BIC                                    & $29626.31$   & $29630.35$   & $29636.59$   \\
Num. obs.                              & $7576$       & $7576$       & $7576$       \\
Num. groups: Individuals               & $1322$       & $1322$       & $1322$       \\
Var: Individuals (Intercept)           & $0.16$       & $0.16$       & $0.16$       \\
Var: Residual                          & $0.49$       & $0.49$       & $0.49$       \\
\hline
\multicolumn{4}{l}{\scriptsize{\parbox[t]{\linewidth}{%
*** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile,\\
subjective social status, age, gender, political position, and inequality perception.}}}
\end{tabular}
}
\label{table:coefficients}
\end{center}

@@

\section*{Appendix}\label{appendix-appendix}
\addcontentsline{toc}{section}{(APPENDIX) Appendix}

@@

\begin{center}
\scalebox{0.765}{
\begin{tabular}{l c c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
\hline
Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.873^{***}$  & $1.741^{***}$  & $1.664^{***}$  & $2.276^{***}$  & $2.277^{***}$  \\
                                            & $(0.026)$      & $(0.038)$      & $(0.090)$      & $(0.092)$      & $(0.092)$      & $(0.105)$      & $(0.138)$      \\
Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                \\
                                            &                &                &                &                &                &                &                \\
\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                \\
\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                \\
\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                \\
\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                \\
                                            & $(0.030)$      &                &                &                &                &                &                \\
\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                \\
                                            & $(0.029)$      &                &                &                &                &                &                \\
Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
Covid: receive state benefits               &                &                & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
                                            &                &                & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
Covid: economy more important than health   &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
                                            &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
Merit: Talent                               &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
                                            &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Merit: Effort                               &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
                                            &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
Inequality perception                       &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
                                            &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
Universitary (Ref.= Less than universitary) &                &                &                &                &                &                & $-0.048$       \\
                                            &                &                &                &                &                &                & $(0.046)$      \\
Household Income                            &                &                &                &                &                &                & $0.001^{*}$    \\
                                            &                &                &                &                &                &                & $(0.000)$      \\
Subjective Social Status                    &                &                &                &                &                &                & $-0.005$       \\
                                            &                &                &                &                &                &                & $(0.012)$      \\
Age (Ref.= 18-29)                           &                &                &                &                &                &                &                \\
                                            &                &                &                &                &                &                &                \\
\quad Age 30-49                             &                &                &                &                &                &                & $-0.033$       \\
                                            &                &                &                &                &                &                & $(0.051)$      \\
\quad Age 50-64                             &                &                &                &                &                &                & $0.009$        \\
                                            &                &                &                &                &                &                & $(0.053)$      \\
\quad Age 65 or more                        &                &                &                &                &                &                & $0.123$        \\
                                            &                &                &                &                &                &                & $(0.066)$      \\
Gender (Ref. Male)                          &                &                &                &                &                &                & $-0.082^{*}$   \\
                                            &                &                &                &                &                &                & $(0.037)$      \\
Pol. pos (Ref.= Left)                       &                &                &                &                &                &                &                \\
                                            &                &                &                &                &                &                &                \\
\quad Center                                &                &                &                &                &                &                & $0.095$        \\
                                            &                &                &                &                &                &                & $(0.051)$      \\
\quad Right                                 &                &                &                &                &                &                & $0.220^{***}$  \\
                                            &                &                &                &                &                &                & $(0.059)$      \\
\quad Does not identify                     &                &                &                &                &                &                & $0.085$        \\
                                            &                &                &                &                &                &                & $(0.047)$      \\
\hline
BIC                                         & $29667.072$    & $29670.297$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
\hline
\multicolumn{8}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
\end{tabular}
\label{table:coefficients}
\end{center}

### 2025-01-21 paper.qmd
add submit archives

---
title: Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
affiliations: 
  - id: 1
    name: Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. This paper is guided by the following research questions: To what extent are such educational disparities justified in society? Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic? And, what are the main factors driving this kind of justification? The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic). Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
---

@@

This document was last modified at {{< modified >}}

and it was last rendered at {{< now >}}

>>>>

---
title: Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
affiliations: 
  - id: 1
    name: Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. This paper is guided by the following research questions: To what extent are such educational disparities justified in society? Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic? And, what are the main factors driving this kind of justification? The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic). Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
---

@@

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```
